Figures to insert in text

### 2018-02-26 Chapters/Chapter02.tex
SDM: canetation + info theory.

Sparse Distributed Memory (SDM) \citep{Kanerva1988} is a mathematical model of long-term memory that has a number of neuroscientific and psychologically plausible dynamics. It has been applied in many different fields, like pattern recognition \citep{norman2003modeling, rao1995natural}, noise reduction \citep{Meng2009}, handwriting recognition \citep{fan1997genetic}, robot automation \citep{Rajesh1998, mendes2008robot}, and so forth. Despite all those applications, there is not a reference implementation which would allow one to replicate the results published in a paper, to check the source code for details, and to improve it. Thus, even though great results have been achieved using SDM, it requires great effort of researchers to improve someone's work.

It is our belief that this tool could bring orders of magnitude more researchers and attention if they were able to use the model, at zero cost, with an easy to use high-level language such as python. Neuroscientists interested in long-term memory storage should not have to worry about high-bandwidth vector parallel computation.  This new tool provides a ready to use system in which experiments can be executed almost as soon as they are designed.

@@

The main contribution of this work is a reference implementation which yields (i) orders of magnitude gains in performance, (ii) has several backends and operations, (iii) has automatic tests, (iv) is cross-platform, and (v) is easily extended to fulfill other research models. Our reference implementation may, hopefully, accelerate research into the model's dynamics and make it easier to readers to replicate any previous results and easily understand the source-code of the model.

SDM memory or an extension of it has been used in many applications, helping to solve specific problems. \citet{Snaider2011} extended SDM to efficiently store sequences of vectors and trees.  \citet{Rajesh1998} used SDM in an autonomous robot. \citet{Meng2009} modified SDM to clean patterns from noisy inputs.

\citet{Linhares2011} showed that SDM respects the limits of short-term memory discussed by \citet{Miller1955} and \citet{Cowan2001}. This is the only work, to the author's knowledge, that explores some of intrinsic properties of SDM.

@@

The framework implements the basic operations in a Sparse Distributed Memory which may be used to create more complex operations. It is developed in C language --- which may be loaded in many platforms and programming languages --- with a wrapper in Python. The Python module makes it easy to create and execute simulations in a Sparse Distributed Memory and works properly in Jupyter Notebook \citep{kluyver2016jupyter}. It works in both Python 2 and Python 3.

@@

Basic operations were grouped in four sets: (i) for bitstrings, (ii) for addresses, (iii) for counters, and (iv) for memories (SDMs). Operations include creating new bitstrings, flipping bits, generating bitstring with a specific distance from a given bitstring, scanning the address space using different algorithms, writing a bitstring to a counter, writing in an SDM, reading from an SDM, and iteratively reading from an SDM until convergence.

@@

\subsection{The distace between two bitstrings}

@@

It is calculated counting the number of ones in the exclusive or (xor) between the bitstrings, i.e., $d(x, y) = \text{number of ones in} x \oplus y$. There are several algorithms to calculate the number of ones \citep{warren2013hacker}, but the performance depends on the processor. So, we have implemented three different algorithms and one may be selected through compiling flags. The default algorithm is to use a builtin instruction from the compiler.

@@

Bitstrings are stored in a contiguous array of 64-bit integers, as shown in Figure \ref{tab:hl-addresses-detail}. Hence, there is some performance improvement when they are accessed \citep{pai2004linux}.

@@

The scan for activated hard-locations is performed in an address space. It returns the indexes of the bitstrings which were inside the circle. Then, each operation uses these index in a different way.

@@

Scanning for the activated hard-locations is a problem similar to well-known problems in computation geometry called ``range reporting in higher dimensions''. In this case, none of the known algorithms is able to solve our problem faster than $O(n)$. The algorithm which seems to best fit in our problem consumes $O(n)$ space and runs in $O(\log^d(n))$ \citep{chazelle1988functional}, which is really slower than $O(n)$ when, for instance, $n=1,000,000$ and $d=1,000$. For a review of the algorithms, see \citet{chan2011orthogonal}.

@@

It is intriguing that none of those algorithms is able to solve our scanning problem. The idea behind those computation geometry algorithms is roughly to split the search space in half each step, which would take $O(\log(n))$ to go through the whole space. But this approach does not work because of the high number of dimensions (i.e., 1,000) and because the hardlocations' addresses are randomly sampled from the ${0, 1}^d$ space. So, each addresses' bit itself splits the hardlocations in half, but it does not split the search space in half since both halves still must be convered by the algorithm. For instance, let's say we have 1,000 dimensions with 1,000,000 hardlocations, and we are scanning within a circle with radius 451, then after checking the first bit we have two cases: (i) for the half with the same first bit, we must keep scanning with radius 451; and (ii) for the half with a different first bit, we must keep scanning with radius 450. The search space has not been split in half because both halves have been covered (and one of them should have been skipped).

@@

When two counters are added in a third counter, there may occur an overflow. Counters may have overflow protection depending on compiling options. By default, there is no overflow check for performance reasons. It is not supposed to be a problem because each counter is a signed 32-bit integer which can store any number between -2,147,483,648 and 2,147,483,647 (and they will not overflow with less than $2^{31}-1$ divided by the number of activated hard-locations). For instance, when $N=1,000$, $H=1,000,000$, and $r=451$, the average number of activated hard-locations is 1,000 and it would require at least one million writes to be possible to overflow a counter.

@@

It is available both Kanerva's read and Chada's read. The latter was implemented according to a generalization proposed by professor Paulo Murilo (personal communication). The generalization has a $z$ parameters, which is the expoent of the counter. In this case, the results are floating point instead of integer, which reduces considerably the performance. When $z=1$, it is exactly as the Kanerva's read. When $z=0$, it is the Chada's read. We also explored how SDM would behave for different values of $z$.

@@

SHOW THE RESULTS --- Kanerva's tables/figures vs Simulation.

The objective here is twofold: (i) a single command will install the framework, and (ii) another single command will run (and display) the desired figures obtained the simulation. This will allow potential users to become familiarized with the system and its underlying code with barely any learning curve.

@@

\begin{figure}[h]
\centering\includegraphics[width=0.5\textwidth]{images02/kanerva-table-7-2-original.png}
\caption{Kanerva's original Figure 7.3 (p. 70) predicting a \textasciitilde 500-bit distance after a point.
\label{kanerva-table-7-2}}
\end{figure}

@@

\begin{figure}[h]
\centering\includegraphics[width=0.5\textwidth]{images02/sdm-10000w-table-7-2.png}
\caption{Results generated by the framework diverging from Kanerva's original Table 7.2. Here we had a 1,000 bit, 1,000,000 hard-location SDM with exactly 10,000 random bitstrings written into it, which was also Kanerva's configuration.
\label{sdm-10000w-table-7-2}}
\end{figure}

But, when we reduced the number of random bitstrings written in the SDM from 10,000 to only 100, the results reflected very well the Kanerva's theoretical expectation (Fig. \ref{sdm-100w-table-7-2}). This result strengthen our theory that the disparities in the computational results are due to the interaction effect of different attractors.

\begin{figure}[h]
\centering\includegraphics[width=0.5\textwidth]{images02/sdm-100w-table-7-2.png}
\caption{Results generated by the framework similar from Kanerva's original Table 7.2. It was a 1,000 bit, 1,000,000 hard-location SDM with exactly 100 random bitstrings written into it.
\label{sdm-100w-table-7-2}}
\end{figure}

@@

\begin{figure}[h]
\centering\includegraphics[width=0.5\textwidth]{images02/sdm-10000w-notX-table-7-2.png}
\caption{This graph shows the interaction effects more clearly.  As we include an opposite bitstring, one can see the accelerating effects towards convergence to both attractors: the origin and the opposite. Here we have the exact same configuration of Figure \ref{sdm-10000w-table-7-2}, with the addition of the single opposite attractor.
\label{sdm-10000w-notX-table-7-2}}
\end{figure}

@@

@@

We propose to provide a new, open-source, cross-platform, highly parallel framework in which researchers may be able to create hypothesis and test them computationally through minimal effort. Though the framework is not well-documented for public release at this time, it has already served as the backbone of Chada's Ph.D. thesis. The single-line command ``pip install sdm'' will install the framework on posix-like systems, and single-line commands will let users test the framework, generate some of the figures from Kanerva's theoretical predictions in their own machines, and --- if interested enough ---, test their own theories and improve the framework in an open-source fashion. It is our belief that such work is a necessary component towards accelerating research in this promising field.

We have generated a Docker image, which makes it even easier to explore the framework. After running the container, a Jupyter Notebook is avaiable with sdm-framework and other tools already installed. The simulations run in this thesis are promptly available to be re-executed and explored. We invite the reader to take a look and explore a little bit.

>>>>

Sparse Distributed Memory (SDM) \citep{Kanerva1988} is a mathematical model of long-term memory that has a number of neuroscientific and psychologically plausible dynamics. It has been applied in many different fields, like pattern recognition \citep{norman2003modeling, rao1995natural}, noise reduction \citep{Meng2009}, handwriting recognition \citep{fan1997genetic}, robot automation \citep{Rajesh1998, mendes2008robot}, and so forth. Despite all those applications, there is not a reference implementation which would allow one to replicate the results published in a paper, to check the source code for details, and to improve it. Thus, even though intriguing results have been achieved using SDM, it requires great effort of researchers to improve someone's work.

It is our belief that such a tool could bring orders of magnitude more researchers and attention if they were able to use the model, at zero cost, with an easy to use high-level language such as python. Neuroscientists interested in long-term memory storage should not have to worry about high-bandwidth vector parallel computation.  This new tool provides a ready to use system in which experiments can be executed almost as soon as they are designed.

@@

The main contribution of this work is a reference implementation which yields (i) orders of magnitude gains in performance, (ii) has several backends and operations, (iii) has automatic tests, (iv) is cross-platform, and (v) is easily extended to fulfill other research models. Our reference implementation may, hopefully, accelerate research into the model's dynamics and make it easier for readers to replicate any previous results and easily understand the source-code of the model.  Moreover, new machanisms and applications are also introduced.

SDM, extensions of it, has been used in many applications. For example, \citet{Snaider2011} extended SDM to efficiently store sequences of vectors and trees.  \citet{Rajesh1998} used SDM in an autonomous robot. \citet{Meng2009} modified SDM to clean patterns from noisy inputs. \citet{Linhares2011} showed that SDM respects the limits of short-term memory discussed by \citet{Miller1955} and \citet{Cowan2001}.

@@

The framework implements the basic operations in a Sparse Distributed Memory which may be used to create more complex operations. It is developed in C language and the OpenCL parallel framework --- which may be loaded in many platforms and programming languages --- with a wrapper in Python. The Python module makes it easy to create and execute simulations in a Sparse Distributed Memory and works properly in Jupyter Notebook \citep{kluyver2016jupyter}. It works in both Python 2 and Python 3.

@@

Basic operations were grouped in four sets: (i) for bitstrings, (ii) for addresses, (iii) for counters, and (iv) for memories (SDMs). Operations include creating new bitstrings, flipping bits, generating a bitstring with a specific distance from a given bitstring, scanning the address space using different algorithms, writing a bitstring to a counter, writing in an SDM, reading from an SDM, and iteratively reading from an SDM until convergence.

@@

\subsection{The distance between two bitstrings}

@@

It is calculated counting the number of ones in the exclusive or (xor) between the bitstrings, i.e., $d(x, y) = \text{number of ones in } x \oplus y$. There are several algorithms to calculate the number of ones \citep{warren2013hacker}, but the performance depends on the processor. So, we have implemented three different algorithms and one may be selected through compiling flags. The default algorithm is to use a builtin instruction from the compiler.

@@

Bitstrings are stored in a contiguous array of 64-bit integers, as shown in Figure \ref{tab:hl-addresses-detail}. Hence, basic pointer arithmetic provides us with performance improvements in their access, as processors realize fechts of contiguous chunks of memory  \citep{pai2004linux}.

@@

The scan for activated hard-locations is performed in an address space. It returns the indexes of the bitstrings which were inside the circle. Then, each operation uses this index in a different way.

@@

Scanning for the activated hard-locations is a problem similar to well-known problems in computational geometry called ``range reporting in higher dimensions''. In this case, none of the known algorithms is able to solve our problem faster than $O(n)$. The algorithm which seems to best fit in our problem consumes $O(n)$ space and runs in $O(\log^d(n))$ \citep{chazelle1988functional}, which is really slower than $O(n)$ when, for instance, $n=1,000,000$ and $d=1,000$. For a review of the algorithms, see \citet{chan2011orthogonal}.

@@

It is intriguing that none of those algorithms is able to solve our scanning problem. The idea behind those computational geometry algorithms is roughly to split the search space in half each step, which would take $O(\log(n))$ to go through the whole space. But this approach does not work because of the high number of dimensions (i.e., 1,000) and because the hardlocations' addresses are randomly sampled from the $\{0, 1\}^d$ space. So, each addresses' bit itself splits the hardlocations in half, but it does not split the search space in half since both halves still must be convered by the algorithm. For instance, let's say we have 1,000 dimensions with 1,000,000 hardlocations, and we are scanning within a circle with radius 451, then after checking the first bit we have two cases: (i) for the half with the same first bit, we must keep scanning with radius 451; and (ii) for the half with a different first bit, we must keep scanning with radius 450. The search space has not been split in half because both halves have been covered (and one of them should have been skipped).

@@

When two counters are added in a third counter, there may occur an overflow. Counters may have overflow protection depending on compiling options. By default, there is no overflow check for performance reasons. It is not supposed to be a problem because each counter is a signed 32-bit integer which can store any number between -2,147,483,648 and 2,147,483,647 (and they will not overflow with less than $2^{31}-1$ divided by the number of activated hard-locations). For instance, when $N=1,000$, $H=1,000,000$, and $r=451$, the average number of activated hard-locations is 1,000 and it would require at least one million writes to be possible to overflow a counter.  Note also that it would be improbable for a hard location dimension to be this biased before the complete saturation of the memory.

@@

The framework is available both Kanerva's read and Chada's read. The latter was implemented according to a generalization proposed by physics professor Paulo Murilo of UFF during the discussion of my master's defense. The generalization brings a parameter $z$, which is the exponent of the counter. In this case, the results are floating point instead of integer, which considerably reduces performance. When $z=1$, it is exactly as the Kanerva's read. When $z=0$, it is the Chada's read. We also explored how SDM would behave for different values of $z$.

@@

The objective here is twofold: (i) a single command will install the framework, and (ii) another single command will run (and display) the desired figures obtained the simulation. This will allow potential users to become familiarized with the system and its underlying code with barely any learning curve beyond scientific python and jupyter notebooks.

@@

\begin{figure}[h]
\centering\includegraphics[width=0.8\textwidth]{images02/kanerva-table-7-2-original.png}
\caption{Kanerva's original Figure 7.3 (p. 70) predicting a \textasciitilde 500-bit distance after a point.
\label{kanerva-table-7-2}}
\end{figure}

@@

\begin{figure}[h]
\centering\includegraphics[width=0.8\textwidth]{images02/sdm-10000w-table-7-2.png}
\caption{Results generated by the framework diverging from Kanerva's original Table 7.2. Here we had a 1,000 bit, 1,000,000 hard-location SDM with exactly 10,000 random bitstrings written into it, which was also Kanerva's configuration.
\label{sdm-10000w-table-7-2}}
\end{figure}

But, when we reduced the number of random bitstrings written in the SDM from 10,000 to only 100, the results reflected very well the Kanerva's theoretical expectation (Fig. \ref{sdm-100w-table-7-2}). This result strengthens our hypothesis that the disparities in the computational results are due to the interaction effect of high numbers of different attractors.

\begin{figure}[h]
\centering\includegraphics[width=0.8\textwidth]{images02/sdm-100w-table-7-2.png}
\caption{Results generated by the framework similar from Kanerva's original Table 7.2. It was a 1,000 bit, 1,000,000 hard-location SDM with exactly 100 random bitstrings written into it.
\label{sdm-100w-table-7-2}}
\end{figure}

@@

\begin{figure}[h]
\centering\includegraphics[width=0.8\textwidth]{images02/sdm-10000w-notX-table-7-2.png}
\caption{This graph shows the interaction effects more clearly.  As we include an opposite bitstring, one can see the accelerating effects towards convergence to both attractors: the origin and the opposite. Here we have the exact same configuration of Figure \ref{sdm-10000w-table-7-2}, with the addition of the single opposite attractor.
\label{sdm-10000w-notX-table-7-2}}
\end{figure}

@@

\section{Results (vi): Information theoretical write operation}

\begin{figure}[h!]
  \centering
  \subfloat[Kanerva's model]{\includegraphics[width=\linewidth]{./images02/new-images/kanerva-15iter-reads.jpeg}}

  \subfloat[Write process weighted by the amount of information contained in the distance between the written bitstring and each hard location]{\includegraphics[width=\linewidth]{./images02/new-images/info-theory-15iter-reads.jpeg}}

  \caption{(a) and (b) show the behavior of the critical distance under Kanerva's model and the information-theoretic one, respectively.}
  \label{fig:info-theory-experiments}
\end{figure}

My advisor, Alexandre Linhares, has proposed another read operation: the weighted reading. In it, the sum of the counter's value is weighted based on the distance between the hard-location's address and the reading address. The logic behind it is to vary the importance of each hard-location inside the circle.  It is only natural that one encodes an item in near hard locations with a stronger signal, and a natural candidate for this signal function is the amount of information contained in the distance between the item and each hard location.  Closer hard locations have lower probabilies and therefore should encode more information. We know that the probability of a distance $d$ is $\binom{N}{d}/2^N$, and, acordingly, we have tested how SDM behaves when write weights are calculated using the amount of information given by each bitstring, i.e., $w(d) = -\log_2 \left( 2^{-N} \binom{N}{d} \right) = N - \log_2 \binom{N}{d}$, as in information theory \citep{cover2012elements}.

The results can be seen in Figure \ref{fig:info-theory-experiments} and seem promising. It seems that the critical distance increases by a number of bits.  Note that 10 additional bits imply an attractor $2^{10}$ of the size of the original. Another point to keep in mind is that, since the modulus of the vectors are not uniform in this approach, that the shape of the attractor may have large asymmetries.

@@

We propose to provide a new, open-source, cross-platform, highly parallel framework in which researchers may be able to create hypotheses and test them computationally through minimal effort. The framework is well-documented for public release at this time (http://sdm-framework.readthedocs.io), it has already served as the backbone of Chada's Ph.D. thesis. The single-line command ``pip install sdm'' will install the framework on posix-like systems, and single-line commands will let users test the framework, generate some of the figures from Kanerva's theoretical predictions in their own machines, and --- if interested enough ---, test their own theories and improve the framework, and the benchmarks used to evaluate the framework, in open-source fashion. It is our belief that such work is a necessary component towards accelerating research in this promising field.

We have generated a Docker image, which makes it even easier to explore the framework. After running the container, a Jupyter Notebook is available with sdm-framework and other tools already installed. The simulations run in this thesis are promptly available to be re-executed and explored. We invite the reader to take a look and explore a little bit.

### 2018-02-26 Chapters/sdm2.tex
SDM: canetation + info theory.

The data and address space belong to binary space and are represented by a sequence of bits, called bitstrings. The distance between two bitstrings is calculated using the Hamming distance. It is defined for two bitstrings of equal length as the number of positions at which the bits are different. For example, $00110_{b}$ and $01100_{b}$ are bitstrings of length 5 and their Hamming distance is 2. One has to be careful when thinking intuitively about distance in SDM because the Hamming distance does not have the same properties that Euclidean distance does.

Every time you are writing or reading in SDM memory, a number of addresses are activated according to their distance to the center address and your data is written in these activated addresses or read from them.  These issues will be addressed in due detail further on, but a major difference from a traditional computer memory is that your data are always stored and retrieved in a multitude of addresses. This way SDM memory has robustness against loss of addresses.

In traditional memory each datum is stored in an address and every time you are looking for a specific datum you have to search through the memory to find out whether you have it. In spite of computer scientists having developed many algorithms to perform fast searches, almost all of them do a precise search. That is, if you have an imprecise clue of what you need, these algorithms will simply fail.

In SDM, the data space is the same as the address space, which amounts to a vectorial, binary space, that is, a $\{0,1\}^{n}$ space. This way, the addresses where the data will be written are the same as the data themselves. For example, the datum $\eta=00101_{b}\in\{0,1\}^{5}$ will be written to the address $\xi=\eta=00101_{b}$. If one chooses a radius of 1, the SDM will activate all addresses one bit away or less from the center address. So, the datum $00101_{b}$ will be written to the addresses $00101_{b}$, $10101_{b}$, $01101_{b}$, $00001_{b}$, $00111_{b}$ and $00100_{b}$. In this case, when one needs to retrieve the data, one could have an imprecise cue at most one bit away from $\eta$, since all addresses one bit away have $\eta$ stored in themselves.  Extending this train of thought for larger dimensions and radius, much more addresses are activated and one can see why SDM is a distributed memory.

@@

The addresses of the $\{0,1\}^{n}$ space grows exponentially with the number of dimensions $n$, i.e. $N=2^{n}$. For $n=100$ we have $N\approx10^{30}$, which is incredibly big when related to a computer memory. Furthermore, \citet{Kanerva1988} suggests $n$ between 100 and 10,000. Recently he has postulated 10,000 as a desirable minimum $N$ (personal communication). To solve the feasibility problem of implementing this memory, Kanerva made a random sample of $\{0,1\}^{n}$, in his work, having $N'$ elements. All these addresses in the sample are called hard-locations. Other elements of $\{0,1\}^{n}$, not in $N'$, are called virtual neurons. This is represented in Figure \ref{fig-hardlocations}.  All properties of read and write operations presented before remain valid, but limited to hard-locations. Kanerva suggests taking a sample of about one million hard-locations.

@@

One interesting view is that neurons in SDM work like pointers. As we write bitstrings in memory, the hard-locations' counters are updated and some bits are flipped. Thus, the activated hard-locations do not necessarily point individually to the bitstring that activated it, but together they point correctly. In other words, the read operation depends on many hard-locations to be successful. This effect is represented in Figure \ref{fig-p1-pointers}: where all hard-locations inside the circle are activated and they, individually, don't point to $\eta$.  But, like vectors, adding them up points to $\eta$. If another datum $\nu$ is written into the memory near $\eta$, the shared hard-locations will have information from both of them and wouldn't point to either.  All hard-locations outside of the circle are also pointing somewhere (possibly other data points). This is not shown, however, in order to keep the picture clean and easily understandable.

@@

Daniel Chada, one member of our research group, proposed another way to read in SDM, in this work called Chada's read. Instead of summing all hard-location counters, each hard-location evaluates its resulting bitstring individually. Then, all resulting bitstring are summed again, and the same rule as Kanerva applies. Table \ref{tab:chada-read} shows an example of Chada's read result bitstring. The counter's values are normalized to 1, for positive ones, or -1, for negative ones, and the original values are the same as in Table \ref{tab:kanerva-read}.

@@

A member of my Master's committee, professor Paulo Murilo, has proposed a generalized reading operation (personal communication), which covers both Kanerva's and Chada's read --- and opens a new venue of potential discoveries. He proposed summing all hard-location counters raised to the power of $z$ while holding the original sign of the counter (positive or negative). Thus, Kanerva's read would be the same as $z=1$, while Chada's would be the same as $z=0$. Hence, we might explore how SDM would behave with other values of $z$, such as 0.5, 2, and 3.

@@

My advisor, professor Alexandre Linhares, has proposed another read operation: the weighted reading. In it, the sum of the counter's value is weighted based on the distance between the hard-location's address and the reading address. The logic behind it is to change the importance of each hard-location inside the circle. We have tested how SDM behaves when the weight is calculated using the information given by each bitstring, i.e., $w(d) = -\log_2 \left( 2^{-N} \binom{N}{d} \right) = N - \log_2 \binom{N}{d}$, as in information theory \citep{cover2012elements}.

>>>>

The data and address space belong to binary space and are represented by a sequence of bits, called bitstrings. The distance between two bitstrings is calculated using the Hamming distance. It is defined for two bitstrings of equal length as the number of positions at which the bits are different. For example, $00110_{b}$ and $01100_{b}$ are bitstrings of length 5 and their Hamming distance is 2. One has to be careful when thinking intuitively about distance in SDM because the Hamming distance does not have the same properties of, say, the Euclidean distance.

Every time write or read in SDM memory activates a number of addresses with close distance.  The data is written in these activated addresses or read from them.  These issues will be addressed in due detail further on, but a major difference from a traditional computer memory is that the data are always stored and retrieved in a multitude of addresses. This way SDM memory has robustness against loss of addresses (e.g., death of a neuron).

In traditional memory, each datum is stored in an address and every look up of a specific datum requires a search through the memory. In spite of computer scientists having developed beautiful algorithms to perform fast searches, almost all of them do a precise search. That is, if you have an imprecise clue of what you need, these algorithms will simply fail.

In SDM, the data space is the same as the address space, which amounts to a vectorial, binary space, that is, a $\{0,1\}^{n}$ space. This way, the addresses where the data will be written are the same as the data themselves. For example, the datum $\eta=00101_{b}\in\{0,1\}^{5}$ will be written to the address $\xi=\eta=00101_{b}$. If one chooses a radius of 1, the SDM will activate all addresses one bit away or less from the center address. So, the datum $00101_{b}$ will be written to the addresses $00101_{b}$, $10101_{b}$, $01101_{b}$, $00001_{b}$, $00111_{b}$, and $00100_{b}$.

In this case, when one needs to retrieve the data, one could have an imprecise cue at most one bit away from $\eta$, since all addresses one bit away have $\eta$ stored in themselves.  Extending this train of thought for larger dimensions and radius, exponential numbers of addresses are activated and one can see why SDM is a distributed memory.

@@

The addresses of the $\{0,1\}^{n}$ space grows exponentially with the number of dimensions $n$, i.e. $N=2^{n}$. For $n=100$ we have $N\approx10^{30}$, which is incredibly large when related to a computer memory. Furthermore, \citet{Kanerva1988} suggests $n$ between 100 and 10,000. Recently he has postulated 10,000 as a desirable minimum $N$ (personal communication). To solve the feasibility problem of implementing this memory, Kanerva made a random sample of $\{0,1\}^{n}$, in his work, having $N'$ elements. All these addresses in the sample are called hard-locations. Other elements of $\{0,1\}^{n}$, not in $N'$, are called virtual neurons. This is represented in Figure \ref{fig-hardlocations}.  All properties of read and write operations presented before remain valid, but limited to hard-locations. Kanerva suggests taking a sample of about one million hard-locations.

@@

One interesting view is that neurons in SDM work like pointers. As we write bitstrings in memory, the hard-locations' counters are updated and some bits are flipped. Thus, the activated hard-locations do not necessarily point individually to the bitstring that activated it, but together they point correctly. In other words, the read operation depends on many hard-locations to be successful. This effect is represented in Figure \ref{fig-p1-pointers}: where all hard-locations inside the circle are activated and they, individually, do not point to $\eta$.  But, like vectors, adding them up points to $\eta$. If another datum $\nu$ is written into the memory near $\eta$, the shared hard-locations will have information from both of them and would not point to either.  All hard-locations outside of the circle are also pointing somewhere (possibly other data points). This is not shown, however, in order to keep the picture clean and easily understandable.

@@

Daniel Chada, one member of our research group, proposed another way to read in SDM, in this work called Chada's read. Instead of summing all hard-location counters, each hard-location evaluates its resulting bitstring individually. Then, all resulting bitstrings are summed again, and the same rule as Kanerva applies. Table \ref{tab:chada-read} shows an example of Chada's read result bitstring. The counter's values are normalized to 1, for positive ones, or -1, for negative ones, and the original values are the same as in Table \ref{tab:kanerva-read}.

@@

A member of my Master's committee, professor Paulo Murilo, has proposed a generalized reading operation (personal communication), which covers both Kanerva's and Chada's read --- and opens a new venue of potential discoveries. He proposed summing all hard-location counters raised to the power of $z$ while holding the original sign of the counter (positive or negative). Thus, Kanerva's read would be the same as $z=1$, while Chada's would be the same as $z=0$. Hence, we will here explore how SDM would behave with other values of $z$, such as 0.5, 2, and 3.

@@
